---
title: "Assignment 1"
author: "wz334"
date: "2/10/2021"
output: html_document
---
Exercise 1: Write an expression to compute the number of seconds in a 365-day year, and execute the expression. 
```{r}
60*60*24*365
```

Exercise 2: Define a workspace object which contains the number of seconds in 365-day year, and display the results.
```{r}
sec<-60*60*24*365
sec
```


Exercise 3: Find the function name for base-10 logarithms, and compute the base-10 logarithm of 10, 100, and 1000 (use the `??` function to search).
```{r}
log10(10)
log10(100)
log10(1000)

```
Exercise 4: What are the arguments of the `rbinom` (random numbers following the binomial distribution) function? Are any default or must all be specified? What is the value returned?
```{r}
?rbinom
rbinom(10,1,0.5)
```

Exercise 5: Display the vector of the number of successes in 24 trials with probability of success `0.2` (20%), this simulation carried out 128 times. 
```{r}
# I am actually not very clear here: why it works when I create sim like this?
sim<-sum(rbinom(24,1,0.2))
for(i in 1:128){
sim[i]<-sum(rbinom(24,1,0.2))
}
```

Exercise 6: Summarize the  result of `rbinom` (previous exercise) with the `table` function.  Which is the most likely result? What is the range of results?
```{r}
table(sim)
```

Exercise 7: Create and display a vector representing latitudes in degrees from $0^\circ$ (equator) to $+90^\circ$ (north pole), in intervals of $5^\circ$. Compute and display their cosines -- recall, the trig functions in R expect arguments in radians. Find and display the maximum cosine.
```{r}
?seq
lat<- seq(0,90,by=5)
lat<-(lat/180)*pi
cos(lat)
max(cos(lat))
```
Exercise 8: Check if the `gstat` package is installed on your system. If not, install it. Load it into the workspace. Display its help and find the `variogram` function. What is its description?
```{r}
#install.packages("gstat")
library(gstat)
#?gstat
#?variogram
```

Exercise 9: Display the classes of the built-in constant `pi` and of the built-in constant `letters`.
```{r}
class(pi)
class(letters)
```

Exercise 10: What is the class of the object returned by the `variogram` function? (Hint: see the heading "Value" in the help text.)
```{r}
library(sp)
data(meuse)
names(meuse)
#It does not work
#lzn.vgm=variogram(log(zinc)~1,data=meuse)
```

Exercise 11: List the datasets in the `gstat` package.
```{r}
data(package="gstat")
```
Exercise 12: Load, summarize, and show the structure of the `oxford` dataset.
```{r}
data(oxford)
summary(oxford)
str(oxford)
```
Exercise 13: load the `women` sample dataset. How many observations (cases) and how many attributes (fields) for each case? What are the column (field) and row names? What is the height of the first-listed woman?
```{r}
data(women)
dim(women)
colnames(women)
women$height[1]
```
Exercise 14: List the factors in the `oxford` dataset.
```{r}
oxford$PROFILE
oxford$PROFILE
```
Exercise 15: Identify the thin trees, defined as those with height/girth ratio more than 1 s.d. above the mean. You will have to define a new field in the dataframe with this ratio, and then use the `mean` and `sd` summary functions, along with a logical expression.
```{r}
data(trees)
trees$Ratio<- trees$Height/trees$Girth
ratio.mean<-mean(trees$Ratio)
ratio.sd<-sd(trees$Ratio)
trees$thintree<-ifelse(trees$Ratio<ratio.mean-ratio.sd,1,0)
```

Exercise 16: Display a histogram of the diamond prices in the `diamonds` dataset.
```{r}
library(ggplot2)
data(diamonds)
hist(diamonds$price)
```
Exercise 17: Write a model to predict tree height from tree girth. How much of the height can be predicted from the girth? 
```{r}
model1<-lm(trees$Height~trees$Girth)
summary(model1)
```
Exercise 18: Write a model to predict tree volume as a linear function of tree height and tree girth, with no interaction.
```{r}
model2<-lm(trees$Volume~trees$Girth+trees$Height)
summary(model2)
```
Exercise 19: Write a function to restrict the values of a vector to the range $0 \ldots 1$. Any values $< 0$ should be replaced with $0$, and any values $>1$ should be replaced with $1$.  Test the function on a vector with elements from $-1.2$ to $+1.2$ in increments of $0.1$ -- see the `seq` "sequence" function.
```{r}
restri<-function(x){
  if(x<0)
  {x=0}
  else if(x>1)
  {x=1}
  return(x) 
}
test<-round(seq(-1.2,1.2,0.1),2)
for(i in 1:length(test)){
  print(restri(test[i]))}
```

